Efficiently stable presentations from error-correcting codes

### 2023-09-18 efficient_stability.tex
added an abstract

\begin{abstract}

\end{abstract}

@@

\begin{definition}[Efficient stability]\label{def:eff-stab-intro}
Let $G = \langle S:R\rangle $ be a finitely presented group, $\mu_S$ a distribution on $S$ and $\mu_R$ a distribution on $R$.\footnote{It is natural, and will always be the case for us, to take $\mu_S$ as the ``marginal'' of $\mu_R$, i.e.\ the distribution on $S$ obtained by sampling $r\sim\mu_R$ and returning a uniformly random $s$ that appears in $r$, when $r$ is considered as a multi-set.} For $\delta:[0,1]\to[0,1]$ such that $\lim_{t\to 0}\delta(t)=0$ and an integer $d\geq 1$ we say that the presentation $G=\langle S:R\rangle$ is $(\delta,\mu_S,\mu_R,d)$-stable if for every $(\eps,\mu_R)$-almost homomorphism $\phi$ from $G$ to $\mU(\C^d)$ there is a unitary representation $\psi$ of $G$ on $\mU_d(\C)$ such that
\[ \Es{s\sim \mu_S} \big\|  \phi(s) - \psi(s) \big \|_f^2 \,\leq\, \delta(\eps)\;.\]
We refer to any function $\delta$ satisfying the above as a ``modulus of stability'' of the presentation.
\end{definition}

@@

\paragraph{Efficient stability.}
Faced with the apparent difficulty of studying the general question, it is time to refine our focus and formulate the question which we \emph{do} address. To start, let us explicitly note that 
while stability has previously (for the more part) been studied as a question about a group, our formulation makes it a question about a \emph{presentation} of the group. In particular it is known~\cite{gowers2017inverse,de2019operator} that for finite groups $G$, which are the only groups we consider, the multiplication table presentation, which has $|G|$ generators and $|G|^2$ relations, is $\delta(\eps)=C\eps$-stable for some constant $C$ that is independent of the group---and this, even for approximate representations in arbitrary tracial von Neumann algebras, as long as one allows the notion of flexible stability referred to above. Now observe that, if $G(h)$ happens to be Abelian, then it provides a \emph{different} presentation~\eqref{eq:gh-intro} of $\Z_2^k$, where $k$ is the dimension of $\ker h$. This presentation could be much smaller --- if $n,k$ are linearly related then it has $O(k)$ generators and relations, as opposed to $2^k$ and $2^{2k}$ respectively. Can such ``efficient'' presentations still be stable? The naive approach, of extending an $\eps$-approximate homomorphism of $G(h)=\langle S:R\rangle$ into an $\eps'$-approximate homomorphism of the multiplication table presentation of $\Z_2^k$, leads to $\eps'=\Omega(k\eps)$ and hence a logarithmic dependence of the modulus of stability on the group size. Is it possible to do better? Glebsky's result for the case of $\Z^2$  mentioned earlier suggests that it is, in some cases, possible; but in what generality?

@@

>>>>

\begin{abstract}
We introduce a notion of \emph{efficient stability} for finite presentations of groups. Informally, a finite presentation using generators $S$ and relations $R$ is \emph{stable} if any map from $S$ to unitaries (say, in some finite von Neumann algebra) that approximately satisfies the relations (in the tracial norm) is close to the restriction of a representation of $G$ to the subset $S$. This notion and variants thereof have been extensively studied in recent years, in part motivated by connections to property testing in computer science. The novelty in our work is the focus on \emph{efficiency} which, informally, places an onus on small presentations (of size, say, logarithmic in the group size) and aims to achieve non-trivial tradeoffs between it and the stability---as, intuitively, more generators and relations place more constraints on the approximate homomorphism, making stability easier to achieve---whereas we wish to minimize the number of variables and constraints.

Equipped with the new definition we give examples. We first show a general method for constructing candidate stable presentations of $\Z_2^k$ from linear error-correcting codes that have an additional property known as \emph{local testability}. While we cannot show that the method always leads to stable presentations, we leverage recent results in the study of nonlocal games in quantum information theory (Ji et al., Discrete Analysis 2021) to show that a specific instantiation of our construction, based on the Reed-Muller family of codes, leads to a stable presentation of $\Z_2^k$ of size $\poly\log(k)$ only. As an application we combine this result with recent work of de la Salle (arXiv:2204.07084) to re-derive a key building block in the recent disproof of Connes' Embedding Problem via complexity theory (Ji et al., arXiv:2001.04383), the Pauli Braiding Test of Natarajan and Vidick (IEEE FOCS'18). 
\end{abstract}

@@

\begin{definition}[Stability]\label{def:eff-stab-intro}
Let $G = \langle S:R\rangle $ be a finitely presented group, $\mu_S$ a distribution on $S$ and $\mu_R$ a distribution on $R$.\footnote{It is natural, and will always be the case for us, to take $\mu_S$ as the ``marginal'' of $\mu_R$, i.e.\ the distribution on $S$ obtained by sampling $r\sim\mu_R$ and returning a uniformly random $s$ that appears in $r$, when $r$ is considered as a multi-set.} For $\delta:[0,1]\to[0,1]$ such that $\lim_{t\to 0}\delta(t)=0$ and an integer $d\geq 1$ we say that the presentation $G=\langle S:R\rangle$ is $(\delta,\mu_S,\mu_R,d)$-stable if for every $(\eps,\mu_R)$-almost homomorphism $\phi$ from $G$ to $\mU(\C^d)$ there is a unitary representation $\psi$ of $G$ on $\mU_d(\C)$ such that
\[ \Es{s\sim \mu_S} \big\|  \phi(s) - \psi(s) \big \|_f^2 \,\leq\, \delta(\eps)\;.\]
We refer to any function $\delta$ satisfying the above as a ``modulus of stability'' of the presentation.
\end{definition}

@@

\paragraph{Efficient stability.}
Faced with the apparent difficulty of studying the general question, it is time to refine our focus and formulate the question which we \emph{do} address. To start, let us explicitly note that 
while stability has previously (for the more part) been studied as a question about a group, our formulation makes it a question about a \emph{presentation} of the group. In particular it is known~\cite{gowers2017inverse,de2019operator} that for finite groups $G$, which are the only groups we consider, the multiplication table presentation, which has $|G|$ generators and $|G|^2$ relations, is $\delta(\eps)=C\eps$-stable for some constant $C$ that is independent of the group---and this, even for approximate representations in arbitrary tracial von Neumann algebras, as long as one allows the notion of flexible stability referred to above. Now observe that, if $G(h)$ happens to be Abelian, then it provides a \emph{different} presentation~\eqref{eq:gh-intro} of $\Z_2^k$, where $k$ is the dimension of $\ker h$. This presentation could be much smaller --- if $n,k$ are linearly related then it has $O(k)$ generators and relations, as opposed to $2^k$ and $2^{2k}$ respectively. Such a gain is essential when one recalls the interpretation of the presentation $G(h)$ as a ``tester'' for $G$ --- the size of the presentation is then directly related to the amount of \emph{randomness} required by the tester (to sample a random relation); and in computer science applications randomness is seen as an essential resource (we discuss this more below, in the context of quantum computing).

Can such ``efficient'' presentations still be stable? The naive approach, of extending an $\eps$-approximate homomorphism of $G(h)=\langle S:R\rangle$ into an $\eps'$-approximate homomorphism of the multiplication table presentation of $\Z_2^k$, leads to $\eps'=\Omega(k\eps)$ and hence a logarithmic dependence of the modulus of stability on the group size. Is it possible to do better? Glebsky's result for the case of $\Z^2$  mentioned earlier suggests that it is, in some cases, possible; but in what generality?

@@

Most of the technical legwork required to prove the theorem is due to prior work in the study of nonlocal games in quantum computing, and in particular~\cite{ji2020quantum} (we explain this connection below). Our contribution in the present work is to make an explicit connection with stability and show how the former results can be ``imported'' to obtain new stability results such as the one stated in our main theorem above.
